---
title: "analysis"
author: "Patrick"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(DT)

knitr::opts_chunk$set(echo = FALSE, warning = F, message = F)
```
# More Entrances than Exits

The shelter system can't keep up with the high number of people entering shelter each month, as the exits to housing are stable given limited resources to help house people in shelter.

```{=html}
<iframe title="Shelter Exits Not Keeping Pace" aria-label="Multiple Lines" id="datawrapper-chart-1CcpR" src="https://datawrapper.dwcdn.net/1CcpR/2/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="282" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();</script>
```

# Flows in and Out of Shelter

The route to housing is slim - people either move around the system. When they do exit, they are likely to exit unknown - more likely meaning they stay homeless than they find housing.

```{=html}
<div class="flourish-embed flourish-sankey" data-src="visualisation/17241243"><script src="https://public.flourish.studio/resources/embed.js"></script></div>
```

# Where do people go?

Overwhelmingly, people are exiting unknown, and increasingly so in recent months, but the picture looks different for families, who are much more likely to get access to housing progrms, than it does for single adults.


```{=html}
<iframe title="A Huge Spike In Unknown Family Exits" aria-label="Multiple Lines" id="datawrapper-chart-ZaBVs" src="https://datawrapper.dwcdn.net/ZaBVs/3/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="737" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();</script>


```

```{=html}
<iframe title="Single Adults Exit Unknown Most of the Time" aria-label="Multiple Lines" id="datawrapper-chart-kly9G" src="https://datawrapper.dwcdn.net/kly9G/1/" scrolling="no" frameborder="0" style="width: 0; min-width: 100% !important; border: none;" height="516" data-external="1"></iframe><script type="text/javascript">!function(){"use strict";window.addEventListener("message",(function(a){if(void 0!==a.data["datawrapper-height"]){var e=document.querySelectorAll("iframe");for(var t in a.data["datawrapper-height"])for(var r=0;r<e.length;r++)if(e[r].contentWindow===a.source){var i=a.data["datawrapper-height"][t]+"px";e[r].style.height=i}}}))}();</script>

```

# Exits

```{r}
exits <- read_csv("data/shelter_exits.csv")
```

## Total exits by month

```{r}
total_exits <- exits %>%
  filter(agency == "DHS") %>% 
  group_by(date) %>% 
  summarize(total_exits = sum(exits, na.rm = T))

total_exits_series <- exits %>%
  filter(agency == "DHS") %>% 
  filter(series %in% c("families_with_children", "total_single_adults", "adult_families")) %>% 
  filter(date >= as.Date("2023-06-01"), date <= as.Date("2023-11-01")) %>% #update this range when re-runnign with new data
  group_by(date) %>% 
  summarize(total_exits = sum(exits, na.rm = T))

total_exits_category_sum <- exits %>%
  filter(agency == "DHS") %>% 
  filter(series %in% c("families_with_children", "total_single_adults", "adult_families")) %>% 
  filter(date >= as.Date("2023-06-01"), date <= as.Date("2023-11-01")) %>% #update this range when re-runnign with new data
  group_by(date, housing_category) %>% 
  summarize(value = sum(exits, na.rm = T))
```

```{r}
total_exits %>% 
  filter(date >= "2023-04-01" & date <= "2023-11-01") %>% 
  ggplot()+
  geom_line(mapping = aes(x= date, y= total_exits))+
  labs(title = "The rate of exits from DHS is stable",
       subtitle = "Exits from DHS shelters by month")
```

```{r}
total_exits %>% 
  filter(date >= "2023-04-01" & date <= "2023-11-01") %>% 
  ggplot()+
  geom_line(mapping = aes(x= date, y= cumsum(total_exits)))+
  labs(title = "The rate of exits from DHS is stable",
       subtitle = "Cumulative exits from DHS shelters by month")
```

## Entrants

```{r}
entrants_dhs <- read_csv(URLencode("https://data.cityofnewyork.us/resource/jiwc-ncpi.csv?$where=category = 'Admission and entrance to DHS-administered shelter facilities' AND facility_or_program_type = 'Unduplicated entrants (households) to DHS-administered facilities'")) %>% 
  mutate(across(.cols = everything(), .fns = ~as.character(str_replace_all(.x, ",|#", "")))) %>% 
  mutate_at(vars(families_with_children:data_period), ~as.numeric(if_else(.x == "<10", "0", .x))) %>% 
  mutate(total_single_adults = if_else(total_single_adults == 0|is.na(total_single_adults), single_men + single_women, total_single_adults),
         total_entrants = total_single_adults + adult_families + families_with_children,
         date = base::as.Date(paste0(data_period, "01"), format = "%Y%m%d"),
)

entrants_dhs_old <- read_csv("https://data.cityofnewyork.us/resource/bdft-9t6c.csv?$query=SELECT%0A%20%20%60agency%60%2C%0A%20%20%60reporting_requirements%60%2C%0A%20%20%60category%60%2C%0A%20%20%60facility_or_program_type%60%2C%0A%20%20%60families_with_children%60%2C%0A%20%20%60adult_families%60%2C%0A%20%20%60total_families%60%2C%0A%20%20%60total_adults_on_families%60%2C%0A%20%20%60total_children%60%2C%0A%20%20%60single_men%60%2C%0A%20%20%60single_women%60%2C%0A%20%20%60total_single_adults%60%2C%0A%20%20%60total_adults%60%2C%0A%20%20%60data_period%60%0AWHERE%0A%20%20caseless_one_of(%0A%20%20%20%20%60category%60%2C%0A%20%20%20%20%22Unduplicated%20Eligible%20Entrants%20to%20DHS-administered%20Facilities%22%0A%20%20)%0AORDER%20BY%20%60data_period%60%20DESC%20NULL%20LAST")
```

```{r include=F}
entrants_dhs %>% select(date, total_entrants) %>% mutate(series = "entrants") %>% rename(total = total_entrants) %>% 
  bind_rows(total_exits %>% rename(total=total_exits) %>% mutate(series = "exits"))

#put together recent entrants and exits
```

```{r}
entrants_exits_mmr <- read_csv("https://data.cityofnewyork.us/resource/rbed-zzin.csv?$query=SELECT%0A%20%20%60agency%60%2C%0A%20%20%60agency_name%60%2C%0A%20%20%60id%60%2C%0A%20%20%60parentid%60%2C%0A%20%20%60service%60%2C%0A%20%20%60goal%60%2C%0A%20%20%60indicator%60%2C%0A%20%20%60retired%60%2C%0A%20%20%60source%60%2C%0A%20%20%60description%60%2C%0A%20%20%60createdon%60%2C%0A%20%20%60desireddirection%60%2C%0A%20%20%60geo%60%2C%0A%20%20%60geotype%60%2C%0A%20%20%60geovalue%60%2C%0A%20%20%60additive%60%2C%0A%20%20%60frequency%60%2C%0A%20%20%60lagtime%60%2C%0A%20%20%60reporting_period%60%2C%0A%20%20%60critical%60%2C%0A%20%20%60measurement_type%60%2C%0A%20%20%60fiscalyear%60%2C%0A%20%20%60valuedate%60%2C%0A%20%20%60acceptedvalue%60%2C%0A%20%20%60acceptedvalueytd%60%2C%0A%20%20%60targetmmr%60%2C%0A%20%20%60targetmmr2%60%0AWHERE%0A%20%20caseless_one_of(%60agency%60%2C%20%22DHS%22)%0A%20%20AND%20caseless_one_of(%0A%20%20%20%20%60indicator%60%2C%0A%20%20%20%20%22Families%20with%20children%20entering%20the%20DHS%20shelter%20services%20system%22%2C%0A%20%20%20%20%22Adult%20families%20entering%20the%20DHS%20shelter%20services%20system%22%2C%0A%20%20%20%20%22Single%20adults%20entering%20the%20DHS%20shelter%20services%20system%22%2C%0A%20%20%20%20%22Adult%20families%20exiting%20to%20permanent%20housing%22%2C%0A%20%20%20%20%22Families%20with%20children%20exiting%20to%20permanent%20housing%22%2C%0A%20%20%20%20%22Single%20adults%20exiting%20to%20permanent%20housing%22%0A%20%20)")

mmr_sum <- entrants_exits_mmr %>% 
  mutate(series = case_when(
    str_detect(indicator, "enter") ~ "entrants",
    str_detect(indicator, "exit") ~ "exits to housing"
  )) %>% 
  group_by(valuedate, series) %>% 
  summarize(count = sum(acceptedvalue, na.rm = T))

write_csv(mmr_sum %>% pivot_wider(names_from = "series", values_from = "count") %>% filter(year(valuedate)>2018), "entrances_exits.csv")
```

```{r}
mmr_sum %>% 
  ggplot()+
  geom_line(aes(x=valuedate, y=count, color=series))+
  labs(title = "People are entering shelter at higher rates than they are exiting to housing",
       subtitle = "DHS shelter entrants and exits to permanent housing",
       caption = "Note: Exits to permanent housing are just one type of exit from shelter")

```

## Total exits by category

The large majority exiting are families with children or total_single adults

```{r}
exits %>% 
  group_by(category, series) %>% 
  summarize(total_exits = sum(exits, na.rm = T)) %>% 
  pivot_wider(names_from = "series", values_from = "total_exits") %>%
  adorn_totals(where = c("row", "col")) %>% 
  DT::datatable(options = list(iDisplayLength = 25))
```

### Total exits by category, proportion of population

```{r}
exits %>% 
  group_by(category, series) %>% 
  summarize(total_exits = sum(exits, na.rm = T)) %>% 
  pivot_wider(names_from = "series", values_from = "total_exits") %>%
  ungroup() %>% 
  mutate_at(vars(adult_families:total_single_adults), ~.x/sum(.x, na.rm=T)) %>% 
  datatable(options = list(iDisplayLength = 25)) %>% 
  formatPercentage(2:5)
```

## Total exits by agency

The huge majority of exits are from DHS

```{r}
exits %>% 
  group_by(agency, series) %>% 
  summarize(total_exits = sum(exits, na.rm = T),) %>% 
  pivot_wider(names_from = "series", values_from = "total_exits") %>%
  DT::datatable()
```

# Monthly exits

Note: DHS adult families are delayed 2 months, single adults delayed 1 month

```{r}
exits %>% 
  group_by(date, series) %>% 
  summarize(total_exits = sum(exits, na.rm = T)) %>% 
  pivot_wider(names_from = "series", values_from = "total_exits") %>%
  adorn_totals(where = c("row", "col")) %>% 
  DT::datatable()

exits %>% 
  filter(agency == "DHS") %>% 
  group_by(date, series) %>% 
  summarize(total_exits = sum(exits, na.rm = T)) %>% 
  ggplot()+
  geom_line(aes(x=date, y = total_exits, color = series))+
  labs(title = "DHS Total Exits by Month",
       subtitle = "all series")

```

## Families with children

Unknown exits are rising in large numbers and now make up most of the exits from shelter

```{r}
dhs_month_category <- exits %>% 
  filter(agency == "DHS") %>% 
  group_by(date, category, series) %>% 
  summarize(exits = sum(exits, na.rm = T)) %>% 
  ungroup() %>% 
  group_by(date, series) %>% 
  mutate(per_exits = exits/sum(exits, na.rm = T)*100)
  #pivot_wider(names_from = "series", values_from = "total_exits") %>%

dhs_month_category %>% 
  filter(series == "families_with_children") %>% 
  ggplot()+
  geom_line(aes(x=date, y = exits, color = category))+
  labs(title = "DHS Total Exits by Type",
       subtitle = "families with children")

dhs_month_category %>% 
  filter(series == "families_with_children") %>% 
  ggplot()+
  geom_line(aes(x=date, y = per_exits, color = category))+
  labs(title = "DHS Percentage of Exits by Type",
       subtitle = "families_with_children")

dhs_month_category %>% 
  filter(series == "families_with_children") %>% 
  arrange(date, desc(exits)) %>% 
  pivot_wider(-per_exits, names_from = "category", values_from = "exits") %>% 
  relocate(date, series, Unknown) %>% 
  write_csv("families_with_children.csv")

dhs_month_category %>% 
  filter(series == "total_single_adults") %>% 
  pivot_wider(-per_exits, names_from = "category", values_from = "exits") %>% 
  write_csv("total_single_adults.csv")


dhs_category <- exits %>% 
  filter(agency == "DHS", series %in% c("families_with_children", "total_single_adults", "adult_families")) %>% 
  group_by(date, category) %>% 
  summarize(exits = sum(exits, na.rm = T)) %>% 
  ungroup() %>% 
  group_by(date) %>% 
  mutate(per_exits = exits/sum(exits, na.rm = T)*100)

dhs_category %>% 
  filter(date >= as.Date("2023-06-01"), date <= as.Date("2023-12-01")) %>% 
  pivot_wider(-per_exits, names_from = "category", values_from = "exits") %>% 
  write_csv("exit_cat_all.csv")

```


## Single Adults

A huge proportion of single adults exiting go to unknown places

```{r}
dhs_month_category %>% 
  filter(series == "total_single_adults") %>% 
  ggplot()+
  geom_area(aes(x=date, y = exits, fill = category))+
  labs(title = "DHS Total Exits by Type",
       subtitle = "total_single_adults")

dhs_month_category %>% 
  filter(series == "total_single_adults") %>% 
  ggplot()+
  geom_line(aes(x=date, y = per_exits, color = category))+
  labs(title = "DHS Percentage of Exits by Type",
       subtitle = "total_single_adults")


```

```{r}
# pull in entrances and population data for Sankey

entrants_pop <- read_csv("https://data.cityofnewyork.us/resource/jiwc-ncpi.csv?$query=SELECT%0A%20%20%60agency%60%2C%0A%20%20%60category%60%2C%0A%20%20%60facility_or_program_type%60%2C%0A%20%20%60families_with_children%60%2C%0A%20%20%60adult_families%60%2C%0A%20%20%60total_families%60%2C%0A%20%20%60total_adults_in_families%60%2C%0A%20%20%60total_children%60%2C%0A%20%20%60single_men%60%2C%0A%20%20%60single_women%60%2C%0A%20%20%60runaway_and_homeless_youth%60%2C%0A%20%20%60anyone_of_another_gender%60%2C%0A%20%20%60total_single_adults%60%2C%0A%20%20%60total_adults%60%2C%0A%20%20%60total%60%2C%0A%20%20%60data_period%60%2C%0A%20%20%60data_period_notes%60%0AWHERE%0A%20%20caseless_one_of(%60agency%60%2C%20%22Department%20of%20Homeless%20Services%20(DHS)%22)%0A%20%20AND%20(caseless_one_of(%0A%20%20%20%20%20%20%20%20%20%20%20%60category%60%2C%0A%20%20%20%20%20%20%20%20%20%20%20%22Admission%20and%20entrance%20to%20DHS-administered%20shelter%20facilities%22%2C%0A%20%20%20%20%20%20%20%20%20%20%20%22Total%20number%20of%20individuals%20utilizing%20city-administered%20facilities%22%0A%20%20%20%20%20%20%20%20%20)%0A%20%20%20%20%20%20%20%20%20AND%20caseless_one_of(%0A%20%20%20%20%20%20%20%20%20%20%20%60facility_or_program_type%60%2C%0A%20%20%20%20%20%20%20%20%20%20%20%22Unduplicated%20entrants%20(households)%20to%20DHS-administered%20facilities%22%2C%0A%20%20%20%20%20%20%20%20%20%20%20%22DHS-administered%20facilities%22%0A%20%20%20%20%20%20%20%20%20))")

entrants_pop_clean <- entrants_pop %>% 
  mutate(across(.cols = everything(), .fns = ~as.character(str_replace_all(.x, ",|#", ""))),
         date = base::as.Date(paste0(data_period, "01"), format = "%Y%m%d"),
         category = case_when(
           facility_or_program_type == "DHS-administered facilities" ~ "population",
           facility_or_program_type == "Unduplicated entrants (households) to DHS-administered facilities" ~ "enter"
         )) %>% 
  mutate_at(vars(families_with_children:data_period), ~as.numeric(if_else(.x == "<10", "0", .x))) %>% 
  mutate_at(vars(category:facility_or_program_type), ~str_trim(str_replace_all(.x, "[0-9]", ""), side = "both")) %>% 
  select(date, category, families_with_children, total_single_adults, adult_families) %>% 
  rowwise() %>% 
  mutate(value = sum(families_with_children, total_single_adults, adult_families, na.rm = T)) %>% 
  pivot_wider(id_cols = c("date"), names_from = category, values_from = value)

flow_sum <- entrants_pop_clean %>% 
  full_join(total_exits_series, by = "date") %>% 
  filter(date >= as.Date("2023-06-01"), date <= as.Date("2023-11-01")) %>% #update this range when re-runnign with new data
  replace_na(list(enter = 0)) %>% 
  mutate(stayed = population - total_exits)

sankey_data <- flow_sum %>% 
  pivot_longer(cols = c("total_exits", "enter", "stayed")) %>% 
  bind_rows(total_exits_category_sum) %>% 
  arrange(desc(date)) %>% 
  group_by(name, housing_category) %>% 
  summarise(value = mean(value, na.rm = T)) %>% 
  mutate(source = case_when(
              name == "enter" ~ "Entered",
              name == "total_exits" ~ "Shelter",
              name == "stayed" ~ "Shelter",
              is.na(name) ~ "Exited"
         ),
         dest = case_when(
              name == "enter" ~ "Shelter",
              name == "total_exits" ~ "Exited",
              name == "stayed" ~ "Stayed in Shelter",
              is.na(name) ~ housing_category
         ),
         step_from = case_when(
              name == "enter" ~ 0,
              name == "total_exits" ~ 1,
              name == "stayed" ~ 1,
              is.na(name) ~ 2
         ),
         step_to = case_when(
              name == "enter" ~ 1,
              name == "total_exits" ~ 2,
              name == "stayed" ~ 2,
              is.na(name) ~ 3
         )
      ) %>% 
  ungroup() %>% 
  select(value, source, dest, step_from, step_to)

write_csv(sankey_data, "sankey.csv")
```

